---
title: "PSTAT 100: Data Science Concepts and Analysis"
subtitle: "Course Schedule and Calendar"
author:
  - name: "Ethan Marzban"
    affiliations:
      - name: "Summer Session A, 2025"
author-title: "Instructor"
affiliation-title: "Quarter"
format: html
page-layout: full
fontsize: 14pt
---

## Course Schedule 

:::{.callout-warning}
## **Note**

_**This page will be updated as we progress through the quarter; please check back regularly for updates!**_
:::


:::{.callout-note}
## **Textbook Abbreviations and Icon Meanings**
-   MDSR = [Modern Data Science with R](https://mdsr-book.github.io/mdsr2e/)
-   IMS = [Introduction to Modern Statistics, 2nd Ed.](https://openintro-ims2.netlify.app/)
-   R4DS = [R for Data Science](https://r4ds.hadley.nz/)
-   ISL = [An Introduction to Statistical Learning with Applications in `R`](https://www.statlearning.com/)
-   AMAW = [All Models are Wrong](http://allmodelsarewrong.github.io/)
-   <i class="fa-solid fa-person-chalkboard"></i> = Lecture
-   <i class="fa-solid fa-laptop"></i> = Lab
-   <i class="fa-solid fa-file"></i> = Paper
:::

```{r setup, include = F}
library(googlesheets4)
library(gt)
library(tidyverse)

gs4_deauth()

google_sheet <- read_sheet("https://docs.google.com/spreadsheets/d/13UosNz75xSGDUQvBV8wzEnkyn0Bd_SHRs7x8Lx1xv7A/edit?usp=sharing")
```

```{r}
#| echo: False

google_sheet %>%
  filter(!is.na(Type)) %>%
  select(-Admin_Notes, -Type, -Day) %>%
  mutate(Date = as.Date(Date)) %>%
  gt() %>%
  fmt_date(Date, date_style = 31) %>%
  sub_missing(columns = c(Week, Reading, Topic, Materials), missing_text = "") %>%
  tab_style(
    style = cell_borders(
      sides = "right",
      color = "#D3D3D3",
      style = "solid"
    ),
    locations = cells_body(
      columns = c(Date, Topic, Reading, Materials)
    )
  ) %>%
  cols_align(
    align = "center", 
    columns = c(Week)
  ) %>%
  cols_align(
    align = "right", 
    columns = c(Date)
  ) %>%
  cols_align(
    align = "left", 
    columns = c(Materials)
  ) %>%
  fmt_markdown(
    columns = c(Week, Reading, Topic, Materials)
  ) %>%
  cols_width(
    Week ~ pct(7),
    Date ~ pct(13), 
    Reading ~ pct(27),
    Topic ~ pct(30),
    Materials ~ pct(23)
  ) %>%
  cols_label_with(fn = function(x) {
      janitor::make_clean_names(x, case = "title") %>%
        toupper() %>%
        stringr::str_replace_all("^|$", "**") %>%
        md()
    }) %>%
  tab_options(
    table.font.size = 16) %>%
  opt_row_striping()
```

## Course Calendar

<iframe src="https://calendar.google.com/calendar/embed?height=600&wkst=1&ctz=America%2FLos_Angeles&bgcolor=%23ffffff&src=dWNzYi5wc3RhdDEwMEBnbWFpbC5jb20&src=YWRkcmVzc2Jvb2sjY29udGFjdHNAZ3JvdXAudi5jYWxlbmRhci5nb29nbGUuY29t&src=ZW4udXNhI2hvbGlkYXlAZ3JvdXAudi5jYWxlbmRhci5nb29nbGUuY29t&color=%23039BE5&color=%2333B679&color=%230B8043" style="border:solid 1px #777" width="900" height="600" frameborder="0" scrolling="no"></iframe>
