---
title: "Solutions"
subtitle: "PSTAT 100: Spring 2024 (Instructor: Ethan P. Marzban)"
author:
  - Ethan P. Marzban
date: "`r Sys.Date()`"
date-format: long
format:
  html:
    page-layout: full
    toc: TRUE
---

```{r setup, echo = F}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F}
## optional code chunk;
## gives shortcut for boldface colored text,
## able to be rendered in both PDF and HTML

bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```

Below are solutions to the labs and homework assignments I gave when I taught this course. Some things to keep in mind:

-   There may be typos, so please proceed with caution
-   There are often many different ways to approach a problem, each of them "correct"! As such, if you work through a problem in a different way than the solutions that's fine - your way is not necessarily wrong.

## Lab Solutions

<a href="../Files/LabSolns/Lab01_Solns.pdf" target="_blank"><button type="button" class="btn btn-primary">Lab01 Solns</button></a> 
<a href="../Files/LabSolns/Lab02_Solns.pdf" target="_blank"><button type="button" class="btn btn-primary">Lab02 Solns</button></a> 
<a href="../Files/LabSolns/Lab03_Solns.pdf" target="_blank"><button type="button" class="btn btn-primary">Lab03 Solns</button></a>
<a href="../Files/LabSolns/Lab04_Solns.pdf" target="_blank"><button type="button" class="btn btn-primary">Lab04 Solns</button></a>
<a href="../Files/LabSolns/Lab05_Solns.pdf" target="_blank"><button type="button" class="btn btn-primary">Lab05 Solns</button></a>



## HW Solutions

<a href="../Files/HWSolns/HW01_Theoretical_Solutions.pdf" target="_blank"><button type="button" class="btn btn-primary">HW01, Theoretical Solutions</button></a> 
<a href="../Files/HWSolns/HW01_Coding_Solutions.pdf" target="_blank"><button type="button" class="btn btn-primary">HW01, Computing Solutions</button></a>
<a href="../Files/HWSolns/HW02_Solns.pdf" target="_blank"><button type="button" class="btn btn-primary">HW02 Solutions</button></a>
