---
title: "PSTAT 100: Lecture 05"
subtitle: "The Geometry of Data"
footer: "PSTAT 100 - Data Science: Concepts and Analysis, Summer 2025 with Ethan P. Marzban"
logo: "Images/logo.svg"
format: 
  clean-revealjs:
    theme: ../slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
    menu:
      side: left
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: Ethan P. Marzban
    affiliations: Department of Statistics and Applied Probability; UCSB <br /> <br />
institute: Summer Session A, 2025
title-slide-attributes:
    data-background-image: "Images/logo.svg"
    data-background-size: "30%"
    data-background-opacity: "0.5"
    data-background-position: 80% 50%
code-annotations: hover
---


$$
\newcommand\R{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\1}{1\!\!1}
\newcommand{\comp}[1]{#1^{\complement}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\SD}{\mathrm{SD}}
\newcommand{\vect}[1]{\vec{\boldsymbol{#1}}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\tmat}[1]{\mathbf{#1}^{\mathsf{T}}}
\newcommand{\vect}[1]{\vec{\boldsymbol{#1}}}
\newcommand{\tvect}[1]{\vec{\boldsymbol{#1}}^{\mathsf{T}}}
\DeclareMathOperator*{\argmax}{\mathrm{arg} \ \max}
$$

```{css echo = F}
.hscroll {
  height: 100%;
  max-height: 600px;
  max-width: 2000px;
  overflow: scroll;
}

.hscroll2 {
  height: 100%;
  max-height: 300px;
  overflow: scroll;
}

.hscroll3 {
  max-width: 2rem;
  overflow: scroll;
}
```

```{r setup, echo = F}
library(tidyverse)
library(countdown)
library(fixest)
library(modelsummary) # Make sure you have >=v2.0.0
library(GGally)
library(ggokabeito)
library(reshape2)
library(pander)
library(gridExtra)
library(cowplot)
library(plotly)
```



## {{< fa maximize >}} Big Data

-   How many of you have heard the term [**big data**]{.alert}?

-   At the most basic level, big data is data that is _big_. But what do we really mean by "big," especially seeing as datasets are comprised by both the _number of observations_ and the _number of variables_?

-   Essentially there are three situations to consider:
    -   A large number of both observations and variables
    -   A much larger number of observations than variables
    -   A much larger number of variables than observations


## {{< fa maximize >}} Big Data

-   With the continued improvement of computers and computing, the first two cases on the previous slide are not of as great concern as they were, say, 10 years ago.

-   However, the third case is still an active area of research.
    -   When we have more variables than observations (or even just an absurdly large number of variables), things start to break down quickly, as we will shortly see.
    
-   So, a natural question arises: given a dataset, are all dimensions (variables?) really necessary to convey all of the necessary information?

-   The answer, in some cases, turns out to be "no."


## {{< fa maximize >}} Big Data

-   We'll spend the better part of two lectures addressing the details behind this answer.

-   There are a few things we'll need to discuss first.
    -   For example, what do we really mean by the "dimension" of a dataset?
    -   What does it mean to "convey necessary information?"
    
-   The very first thing we'll do today is to establish a more mathematical framework for discussing data.
    -   This is somewhat in contrast to the "data science" view we took last week, breaking data into its semantics and structure.


-   First, let's take a brief interlude to talk about some [**numerical summaries**]{.alert} of data.


# Interlude: Summary Statistics {background-color="black" background-image="https://media3.giphy.com/media/v1.Y2lkPTc5MGI3NjExaTNhdWN0eTE4N3h1NzBxZml3M2xwMjJyaTF3Z2o2MTk5c3V5cTJjbyZlcD12MV9pbnRlcm5hbF9naWZfYnlfaWQmY3Q9Zw/dsKnRuALlWsZG/giphy.gif" background-size="1500px"}


## {{< fa list >}} Summary Statistics
### Central Tendency

    
-   Given a list **`x`** = (`x`~1~, ..., `x`~_n_~) of _n_ numbers, there are a series of numerical summaries we can provide.
    -   I suspect you will all have seen these concepts before, so I'll go through them relatively quickly.
    
-   First question: what is the "center" (or "most typical value") of `x`? (Relates to [**measures of central tendencey**]{.alert})
    
-   Two main answers: [**median**]{.alert} and [**mean**]{.alert}.

-   To find the **median**, line up the data in ascending order and then tick off the first and last elements; repeat this process until you are either left with one value (the median), or two values (in which case you add these two values and divide by 2 to obtain the median).


## {{< fa list >}} Summary Statistics
### Central Tendency

-   For example, the median of the set $\{-1, 4, 5, 6, 10\}$ is 5 and the median of the set $\{-1, 4, 5, 6, 10, 12\}$ is 5.5 (I encourage you to do this computation on your own, as practice).

-   The definition of the **sample mean** is
$$ \overline{x}_n := \frac{1}{n} \sum_{i=1}^{n} x_i $$

-   So, for example, the mean of $\{-1, 4, 5, 6, 10\}$ is 4.8, which we calculated by summing up all five elements and then dividing by 5.

## {{< fa list >}} Summary Statistics
### Central Tendency

::: {.nonincremental}
-   One way to interpret the mean is as follows: imagine placing a weight of unit 1 at each point _x_~_i_~. Then, imagine finding the point such that a fulcrum placed at this point would result in a perfectly balanced system. The location of this point is the sample mean $\overline{x}_n$.
:::

::: {style="text-align:center"}
![](Images/mean.svg){width="75%"}
:::

-   In `R`, we use `mean()` and `median()` to compute the mean and median (respectively) of a set of numbers, as we saw in Lab01.

## {{< fa list >}} Summary Statistics
### Variability

-   To express how "spread out," or "variable," $\vect{x}$ is, there are three common measures:
    -   [**Range**]{.alert}: $\max\{\vect{x}\} - \min\{\vect{x}\}$
    -   [**Sample Variance:**]{.alert} $\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \overline{x}_n)^2$
        -   The [**sample standard deviation**]{.alert} is just the square root of the sample variance.
    -   [**Inter-Quartile Range**]{.alert} (IQR): $Q_3 - Q_1$

-   The variance is perhaps most commonly used, as it has a simple interpretation as the "average distance from the center."

    -   In `R`, these are computed using `diff(range())`, `var()`, and `IQR()`, respectively.

## {{< fa pencil >}} Your Turn!

::: {.callout-tip}
## **Your Turn!**

Consider the list of numbers `x` = $(1.1, \ 2.4, \ 5.6, \ 7.8, \ 100.1)$. 


:::{.nonincremental}
a)    Calculate both the mean and median of `x`. Which do you think is a "better" description of the central value of `x`?

b)    Calculate the range, [**standard deviation**]{.alert} (square root of variance), and IQR of `x`. Which do you think is a "better" description of the spread of `x`?

:::

:::

```{r}
#| echo: False
#| eval: True

countdown(minutes = 4L, font_size = "4rem")
```


# Back to Data {background-color="black" background-image="https://media4.giphy.com/media/v1.Y2lkPTc5MGI3NjExZjJkZmIyaTdmZGpweG0xMHEwdGtra3R3Z2FoM3R3cm5xMGkzOTRoeSZlcD12MV9pbnRlcm5hbF9naWZfYnlfaWQmY3Q9Zw/QpVUMRUJGokfqXyfa1/giphy.gif" background-size="1500px"}


## {{< fa table >}} Data Matrix

-   Let's now consider a dataset comprised of _n_ observations across _p_ numerical variables. 

:::: {.columns}

::: {.column width="50%"}
::: {.fragment}
|**Name** | **Height (in)** | **Weight (lbs)** |
|:-------:|:---------------:|:----------------:|
| Alex | 61.5 | 130.3 |
| Biyonka | 72.4 | 180.6 |
| Catherine | 58.4 | 86.7 |
:::
:::

::: {.column width="50%"}

-   As an example, consider the dataset to the left with three observations on two numerical variables (height and weight).
:::

::::

-   Stripping away the `Name` column and any extraneous formatting information (cell borders, column titles, etc.), what _mathematical_ object are we left with?
    -   That is, what mathematical object is a grid of numbers?

## {{< fa table >}} Data Matrix

![](Images/data_two_views_1.svg)

## {{< fa table >}} Data Matrix

![](Images/data_two_views_2.svg)


## {{< fa table >}} Data Matrix

-   More generally, consider a dataset with _n_ observations and _p_ variables. 

-   The [**data matrix**]{.alert} is the (_n_ × _p_) matrix $\mat{X} = \{x_{ij}\}$ such that _x_~_ij_~ is the _i_^th^ observation on the _j_^th^ variable.

-   As indicated in the diagram on the previous slide, there are two ways to think of a matrix: we can call these "row-wise" and "column-wise" viewpoints
    -   In the "column-wise" viewpoint, we think of $\mat{X}$ as being comprised of _p_ vectors of length _n_
    -   In the "row-wise" viewpoint, we think of $\mat{X}$ as being comprised of _n_ [transposed]{.underline} vectors of length _p_ 
    -   Essentially, these two viewpoints relate the question: is our data a collection of _observations_, or a collection of _variables_?


## {{< fa table >}} Data Matrix {style="font-size:30px"}
### Two Viewpoints

-   For illustrative purposes, consider again our mock height-weight dataset from before:
$$ \mat{X} = \begin{pmatrix} 61.5 & 130.3 \\ 72.4 & 180.6 \\ 58.4 & 86.7 \\ \end{pmatrix} $$

-   The row-wise viewpoint says: our dataset is comprised of three transposed vectors in $\R^2$:
$$ \left\{ \begin{pmatrix} 61.5 \\ 130.3 \\ \end{pmatrix}, \  \begin{pmatrix} 72.5 \\ 180.6 \\ \end{pmatrix} , \  \begin{pmatrix} 58.4 \\ 86.7 \\ \end{pmatrix}  \right\} $$
    -   Each element represents an _individual_.
    
    
## {{< fa table >}} Data Matrix
### Two Viewpoints

-   The column-wise viewpoint says: our dataset is comprised of two vectors in $\R^3$:
$$ \left\{ \begin{pmatrix} 61.5 \\ 72.4 \\ 58.4 \\ \end{pmatrix}, \ \begin{pmatrix} 130.3 \\ 180.6 \\ 86.7 \\ \end{pmatrix} \right\} $$
    -   Each element now represents a _variable_.

-   Let's see what happens if we "plot" our data under these two viewpoints.

## {{< fa vector-square >}} Geometry of Data
### An Example

```{r}
data.frame(var1 = c(61.5, 72.4, 58.4), var2 = c(130.3, 180.6, 86.7), 
           ind = c("Alex", "Biyonka", "Catherine")) %>%
  plot_ly(
    x = ~var1,
    y = ~var2,
    text = ~paste(ind, 
                  "<br>Height:", var1,
                  "<br>Weight:", var2)
  ) %>%
  add_markers(
    marker = list(size = 20)
  ) %>%
  layout(
    title = list(
      text = "Row-Wise Viewpoint",
      font = list(size = 24)
    ),
    xaxis = list(title = list(text ='var1 (height)')),
    yaxis = list(title = list(text ='var2 (weight)')),
    margin = list(t = 75, b = 75)
  )
```

::: {.fragment}
**Cloud of Individuals:** each point represents an individual in the dataset.
:::


## {{< fa vector-square >}} Geometry of Data
### An Example

```{r}
data.frame(obs1 = c(61.5, 130.3), obs2 = c(72.4, 180.6), obs3 = c(58.4, 86.7), 
           ind = c("height", "weight")) %>%
  plot_ly(
    x = ~obs1,
    y = ~obs2,
    z = ~obs3,
    text = ~paste("Variable", ind, 
                  "<br>Individual 1:", obs1,
                  "<br>Individual 2:", obs2,
                  "<br>Individual 3:", obs3)
  ) %>%
  add_markers(
    marker = list(size = 20)
  ) %>%
  layout(
    title = list(
      text = "Column-Wise Viewpoint",
      font = list(size = 24)
    ),
    margin = list(t = 75, b = 75)
  )
```


::: {.fragment}
**Cloud of Variables:** each point represents the direction of a variable in the dataset ("how can the variable be described based on the individuals?").
:::


## {{< fa table >}} Data Matrix
### Two Perspectives

-   There isn't necessarily one viewpoint (row-wise or column-wise) that is always "better" than the other.
    -   Indeed, it's often useful to be able to switch between the two viewpoints.
    
-   When dealing with numerical considerations, however, the column-wise viewpoint is often preferred.

-   Firstly, all values in a column of a data matrix will be of the same _units_; this is not the case of the values across a _row_ of the data matrix.

-   Secondly, the column-wise viewpoint allows us to compare _variables_ as opposed to comparing _units_.

-   So, let's stick with the column-wise viewpoint for now.


<!-- # Geometry of Data {background-color="black" background-image="https://media1.giphy.com/media/v1.Y2lkPTc5MGI3NjExMW5lanV3ZmhscDNpNTAxenNkcDZoaTl6dDN0czZ3dzlibWsydjZyeiZlcD12MV9pbnRlcm5hbF9naWZfYnlfaWQmY3Q9Zw/j0Oieu4Sa2YFfWIl5L/giphy.gif" background-size="fill"} -->

## {{< fa vector-square >}} Geometry of Data {style="font-size:30px"}

-   It turns out that many of the summary statistics we talked about earlier today have very nice correspondences with quantities from linear algebra.

-   As an example, consider a vector $\vect{x} = (x_1, \cdots, x_n)^{\mathsf{T}}$ that represents a column from a particular data matrix (i.e. a variable in the cloud of variables). Further suppose that $\overline{x}_n = 0$ (i.e. that the data is [**mean-centered**]{.alert})

-   We then have that
$$ \| \vect{x} \| := \sum_{i=1}^{n} x_{i}^2 \  \stackrel{(\overline{x}_n = 0)}{=} \  \sum_{i=1}^{n} (x_i - \overline{x}_n)^2 $$ 
    -   Which summary statistic does this look like?

## {{< fa vector-square >}} Geometry of Data {style="font-size:30px"}

-   For two vectors $\vect{x} = (x_1, \cdots, x_k)^{\mathsf{T}}$ and $\vect{y} = (y_1, \cdots, y_k)^{\mathsf{T}}$, their dot product 
$$ \langle \vect{x}, \vect{y} \rangle := \vect{x} \cdot \vect{y} = \sum_{i=1}^{n} x_i y_i $$ 
can be interpreted in terms of the _sample covariance_ between two sets of observations.

-   The mean is related to the inner product between $\vect{x}$ and the [**unity vector**]{.alert}:
$$ \langle \vect{x} , \vect{1} \rangle = \begin{pmatrix} x_1 & \cdots & x_n \\ \end{pmatrix} \begin{pmatrix} 1 \\ \vdots \\ 1 \\ \end{pmatrix} = \sum_{i=1}^{n} x_i$$

## {{< fa vector-square >}} Geometry of Data

-   So, again, we see that many of our familiar "statistical" quantities have direct correspondences with Linear Algebra quantities - this is one of the reasons Linear Algebra is so important in Statistics!

-   Furthermore, this connection allows us to (in the column-wise viewpoint) obtain summaries of our variables by _performing familiar geometric operations_.
    -   Want to know how correlated two variables in a dataset are? Take the inner product of their observations!
    -   Want to know the variability of a particular variable? Find the norm of the vector of observations!

-   Speaking of Linear Algebra (and slowly turning our attention back to the initial question from the start of today's lecture), let's discuss the "dimensionality" of a dataset.

## {{< fa cubes >}} Dimensionality
### Different Perspectives

\

:::: {.columns}
::: {.column width="50%"}

::: {.fragment}
| Height (in) | Weight (lbs) |
|:-----------:|:------------:|
| 61.5 | 130.3 |
| 72.4 | 180.6 |
| 58.4 | 86.7  |
:::

:::{.fragment}
This dataset seems to have 2 dimensions
:::

:::

::: {.column width="50%"}
::: {.fragment}
| Height (in) | Height (cm) |
|:-----------:|:------------:|
| 61.5 |  156.21
| 72.4 | 183.90 |
| 58.4 | 148.34 |
:::

:::{.fragment}
How many dimensions does this data have?
:::

:::{.fragment}
(How many variables contribute new information?)
:::

:::
::::


## {{< fa cubes >}} Dimensionality
### Different Perspectives

-   From a "data science" perspective, we can think of "dimensionality" (in the column-wise viewpoint) as the number of variables that provide new information.
    -   For instance, in the dataset on the right of the previous slide, we really only have _one_ variable (`height`) - just because we've written it down twice doesn't mean we've gained any new information.

-   This is essentially in line with the "Linear Algebra viewpoint" of dimensionality - the _dimension_ of a matrix is just its rank.
    -   In the column-wise viewpoint, we can think of dimension as just the _column_ rank - i.e. the number of linearly independent columns.
    
-   But, there are a few differences:

## {{< fa cubes >}} Dimensionality
### Different Perspectives

:::: {.columns}

::: {.column width="33%"}
::: {.fragment}
```{r}
#| echo: false
#| fig.height: 12

data.frame(x = seq(1, 5, length = 20), y = seq(1, 5, length = 20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 8) +
  theme_minimal(base_size = 60)
```
:::

::: {.fragment}
How many dimensions does this data have?
:::
:::

::: {.column width="33%"}
::: {.fragment}
```{r}
#| echo: false
#| fig.height: 12

x <- seq(1, 5, length = 20)
y <- seq(1, 5, length = 20)

y[12] <- 4.5

data.frame(x = x, y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 8) +
  theme_minimal(base_size = 60)
```
:::

::: {.fragment}
How many dimensions does this data have?
:::
:::

::: {.column width="33%"}
::: {.fragment}
```{r}
#| echo: false
#| fig.height: 12

x <- seq(1, 5, length = 20)
y <- x^2

data.frame(x = x, y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 8) +
  theme_minimal(base_size = 60)
```
:::

::: {.fragment}
How many dimensions does this data have?
:::
:::

::::


## {{< fa cubes >}} Dimensionality
### Different Perspectives

:::: {.columns}

::: {.column width="33%"}
```{r}
#| echo: false
#| fig.height: 12

data.frame(x = seq(1, 5, length = 20), y = seq(1, 5, length = 20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 8) +
  theme_minimal(base_size = 60)
```

::: {.fragment style="text-align:center"}
Only 1!
:::
:::

::: {.column width="33%"}
```{r}
#| echo: false
#| fig.height: 12

x <- seq(1, 5, length = 20)
y <- seq(1, 5, length = 20)

y[12] <- 4.5

data.frame(x = x, y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 8) +
  theme_minimal(base_size = 60)
```

::: {.fragment style="text-align:center"}
**Linear Algebra:** 2 <br /> **Data Science**: 1
:::
:::

::: {.column width="33%"}
```{r}
#| echo: false
#| fig.height: 12

x <- seq(1, 5, length = 20)
y <- x^2

data.frame(x = x, y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 8) +
  theme_minimal(base_size = 60)
```

::: {.fragment style="text-align:center"}
**Linear Algebra:** 2 <br /> **Data Science**: 1
:::
:::

::::


## {{< fa cubes >}} Dimensionality
### Different Perspectives

-   Even though the column rank of the data matrix corresponding to the middle plot would be 2, the resulting matrix would be _very nearly_ singular (noninvertible) and hence we are more likely to say it is of dimension 1.
    -   Additionally, as we saw in the final panel, if two variables are perfectly related (albeit in a nonlinear fashion), we'd be tempted to say they contain the same information and are, in essence, the same variable repeated twice.
        -   For example, writing down `height` and then `height squared` doesn't really give two new variables!

## {{< fa maximize >}} Big Data (Again)

-   Alright, so let's return to our "big data problem" from the start of today's lecture.

-   Already we can perhaps see some justification for why I said, in certain cases, not all variables are needed to convey the full story of a dataset.

:::: {.columns}

::: {.column width="50%"}
::: {.fragment}
| Height (in) | Height (cm) |
|:-----------:|:------------:|
| 61.5 |  156.21
| 72.4 | 183.90 |
| 58.4 | 148.34 |
:::
:::

::: {.column width="50%"}
-   The second column is providing redunant information.
    -   Again, this is really a one-dimensional dataset!
:::

::::

-   Now, admittedly, this is a bit of an "extreme" situation - most datasets don't just have copies of one variable.
    -   So, how can we figure out which variables (if any) provide redundant information?


# Dimension Reduction {background-color="black" background-image="https://media4.giphy.com/media/v1.Y2lkPTc5MGI3NjExYWQxZnN0NHg4NTBqdHhzN3RyMjRkODF3eWhkNTVhNHR6ejE1ejlyaSZlcD12MV9pbnRlcm5hbF9naWZfYnlfaWQmY3Q9Zw/S9heJFkO9ay9fU4HeS/giphy.gif" background-size="2000px"}

## {{< fa bullseye >}} The Goal

::: {.fragment}
::: {.callout-tip}
## **Goal:** [**Dimension Reduction**]{.alert}

Reduce the dimension of a dataset with as little loss of information as possible.
:::
:::

-   Two questions arise: 
    1)    How do we "reduce dimensions"?
    2)    What do we mean by "little loss of information?"
    
-   Indeed, one (admittedly crude) form of dimension reduction is to simply remove one or more columns from a dataset!

-   But, a slightly more creative approach involves leveraging _projections_. 
    -   Recall that we can (geometrically) think of projections as "shadows" cast by a set of points (vectors) onto a subspace.



## {{< fa video >}} Projecting into Smaller Dimensions

![](Images/step0.svg)


## {{< fa video >}} Projecting into Smaller Dimensions

![](Images/step1_1.svg)


## {{< fa video >}} Projecting into Smaller Dimensions

![](Images/step1_2.svg)



## {{< fa video >}} Projecting into Smaller Dimensions

![](Images/step2_1.svg)


## {{< fa video >}} Projecting into Smaller Dimensions

![](Images/step2_2.svg)



## {{< fa video >}} Projecting into Smaller Dimensions

![](Images/step3_1.svg)


## {{< fa video >}} Projecting into Smaller Dimensions

![](Images/step3_2.svg)



## {{< fa video >}} Projecting into Smaller Dimensions

![](Images/two_proj.svg)

::: {.fragment}
Seems like we might want to preserve as much [variance]{.underline} as possible!
:::

## {{< fa bullseye >}} The Goal, Revisited

::: {.callout-tip}
## **Goal**

Identify the directions (in the cloud of variables) along which there is maximal variance. Then, project onto a subspace spanned by these directions to obtain a low-dimensional representation of the data.
:::

-   We'll set up some of the math today, and continue our discussion tomorrow. 

-   First, let's work toward a more specific goal:

::: {.fragment}
::: {.callout-tip}
## **Goal**

Identify the vector $\vect{v}$ such that $\mat{X} \vect{v}$ (the column mean-centered data projected onto $\vect{v}$) has maximum variance (when compared to all other such vectors).
:::
:::

## {{< fa minimize >}} PCA: A First Pass
### Direction of Maximal Variance

-   Crucially, we'll assume the columns of $\vect{X}$ have mean zero. 
    -   In `R`, we can accomplish this using the `scale()` function.
    
-   Remember how to express the variance of $\mat{X} \vect{v}$ in terms of standard linear algebra operations?
    -   Squared-norm!

-   The variance of $\mat{X} \vect{v}$ is proportional to
$$ \| \mat{X} \vect{v} \|^2 =  (\mat{X} \vect{v})^{\mathsf{T}}(\mat{X} \vect{v}) = \tvect{v} \tmat{X} \mat{X} \vect{v} $$
    -   So, the direction $\vect{v}$ of maximal variance is the one for which $\tvect{v} \tmat{X} \mat{X} \vect{v}$ is maximized. 

## {{< fa minimize >}} PCA: A First Pass
### Direction of Maximal Variance

-   This maximization problem, however, is ill-posed without further constraints. 
    -   Specifically, if $\vect{v}$ is a direction of maximal variance, so too will $c \vect{v}$ be for any constant $c$.

-   So, let's add the following constraint: $\vect{v}$ must be of unit-length.

::: {.fragment}
::: {.callout-tip}
## **Goal**

Identify the unit vector $\vect{v}$ such that $\mat{X} \vect{v}$ (the column mean-centered data projected onto $\vect{v}$) has maximum variance.
:::
:::

::: {.fragment style="font-size:28px}
$$ \argmax_{\vect{v}} \left\{ \tvect{v} \tmat{X} \mat{X} \vect{v} \right\} \quad \text{s.t.} \quad \tvect{v} \vect{v} = 1 $$
:::

## {{< fa minimize >}} PCA: A First Pass
### Direction of Maximal Variance

-   Constrained optimization problems like this are most often solved using [**Lagrange Multipliers**]{.alert}, which we will discuss further in a future lab.
    -   If you'd like a primer, I encourage you to check out [this](https://www.youtube.com/watch?v=5A39Ht9Wcu0) video.
    -   Again, I don't expect you to follow _every_ step of the following computation yet!

-   We first construct the [**lagrangian**]{.alert} $\mathcal{L}(\vect{v}, \lambda) := \tvect{v} \tmat{X} \mat{X} \vect{v} - \lambda \tvect{v} \vect{v}$. 

-   Then we differentiate wrt. $\vect{v}$ and set equal to zero:
$$ \tmat{X} \mat{X} \vect{v} - \lambda \vect{v} = 0 $$
or, in other words, $(\tmat{X} \mat{X}) \vect{v} = \lambda \vect{v}$.

## {{< fa minimize >}} PCA: A First Pass
### Direction of Maximal Variance


::: {.callout-tip}
## **Goal**

Identify the vector $\vect{v}$ such that $\mat{X} \vect{v}$ (the column mean-centered data projected onto $\vect{v}$) has maximum variance.
:::


::: {.fragment}
::: {.callout-important}
## **Result**

The vector $\vect{v}$ is given by an **eigenvector** of $\tmat{X} \mat{X}$ with eigenvalue $\lambda$.
:::
:::

-   The variance of the projected data along $\vect{v}$ is proportional to $\tvect{v} \tmat{X} \mat{X} \vect{v}$.
    -   Since $\vect{v}$ is an eigenvalue of $\tmat{X} \mat{X}$, this simplifies to $\vect{v}(\lambda \vect{v}) = \lambda \tvect{v} \vect{v}$
    -   We assumed $\vect{v}$ is of unit length, meaning the variance is proportional to the associated eigen*value* $\lambda$.


## {{< fa minimize >}} PCA: A First Pass
### Directions of Maximal Variance

-   In other words, the eigen*vectors* $\vect{v}$ give the _directions_ of maximal variance, and the eigen*values* $\lambda$ give the _amount_ of variance the projected data will have (all up to proportionality constants).

-   So, the eigenvector $\vect{v}_1$ associated with the largest eigenvalue is the direction with the largest variance; the eigenvector $\vect{v}_2$ associated with the second-largest eigenvalue is the direction with the second-largest variance; etc.

-   Pretty neat, huh?
    -   Seems like as good a place as any to leave off for today.

## {{< fa forward-fast >}} Next Time

-   We will continue discussing the details of PCA.
    -   For example, now that we have the directions of maximal variance - what do we do?
    -   How many dimensions are enough to capture a "good" lower-dimensional approximation of our data?
    -   How do we do this in `R`?

-   **Reminder:** ICA01 is coming up this Thursday (July 3, 2025).
    -   An information document has been posted to the course website; please read it fully!
    
-   **Second Reminder:** group formation for the Mid-Quarter Project is due **tomorrow** (Tuesday, July 1) by 11:59pm.